Liveness-Based Garbage Collection for Lazy Languages

### 2016-04-26 paper/paper.tex
minor updates after discussion with Amey

\begin{figure}[t]\footnotesize
\renewcommand{\arraystretch}{0.9}
\begin{eqnarray*}
   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\,\,\, e_\mainpgm
   \hspace{5em} \,\,\,\,\,\,\,\,\,\; \mbox{\em --- program}\\
    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
\,\,x_n)\,\,
    e)
    \hspace{0.2em} \ \ \ \ \ \ \ \ \  \mbox{\em --- function def} \\
e \in \mathit{Expr} & \!\!\!::=\!\!\! &
\left\{\begin{array}{@{}ll@{\hspace{1em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
binding} \\
       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{App} & \!\!\!::=\!\!\!  &
\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
       (\CAR\,\, x) & \\
       (\CDR\,\, x) & \mbox{\em --- selectors} \\
       (\NULLQ\,\, x) & \\
       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester/generic arithmetic} \\
%%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
    \end{array}\right.
\end{eqnarray*}
\vspace*{-6pt}
  \caption{The syntax of our language}\label{fig:lang-syntax}
%\figrule
\normalsize
\vspace*{-6pt}
\end{figure}

@@

Closure  variables  get  their  liveness values  through  a  chain  of
dependences beginning  at a  variable at an  evaluation point.   As an
example,  consider the program  in Figure~\ref{fig:lv-closure} again.
The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
expression $(+~\py~1)$.  $\py$ in turn depends on the closure variable
$\px_{\mathit{\pi}_1}$  through $(\length~\px_{\mathit{\pi}_1})$.   We
denote this  chain of  dependences as  $\lbrack \psi_3\!\!\!:\!\!\!\pz
\leftarrow            (+~\py~1),\,           \py            \leftarrow
(\length~\px_{\mathit{\pi}_1})\rbrack$.    Indeed,   the   chains   of
closures  in   the  heap  are    runtime  representations   of  these
dependences.   Since  $\pz$  is  evaluated  at  $\psi_3$  due  to  the
expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
$\px_{\pi_1}$.  Other dependence chains which result in a liveness for
$\px_{\pi_1}$    are    $\lbrack    \psi_1\!\!\!:\!\!\py    \leftarrow
(\length~\px_{\pi_1})\rbrack$   and  $\lbrack   \psi_2\!\!\!:\!\!\!\pw
\leftarrow          (/~\pu~\py),\,\,           \py          \leftarrow
(\length~\px_{\pi_1})\rbrack$.   The  liveness analysis  described  in
this section declares  the liveness of $\px_{\pi_1}$ to be  a union of
the liveness arising  from these dependence chains.  To be  safe, a GC
during evaluation  of $\py$ at  $\psi_1$ has  to use this  liveness to
copy the heap starting from  $\px_{\pi_1}$.  However, notice that if a
GC  takes place  while evaluating  $\pz$  at $\psi_3$,  it can  safely
consider only the liveness arising from the dependence chain\linebreak
$\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow (+~\py~1),\, \py \leftarrow
(\length~\px_{\pi_1})\rbrack$.    The    garbage   collection   scheme
described in Section~\ref{sec:GC-scheme} uses a generalization of this
observation  to  dynamically  select   an  evaluation  point  specific
liveness in order to collect more garbage.

@@

Based on  the above considerations,  we restrict the  possible garbage
collection points in a function body to the following:
\begin{enumerate}
\item We  maintain static estimates  of the memory required  to create
  the closures for each function body.  On entering a function, if the
  available memory is  less than this requirement for  the function, a
  GC is triggered.
\item Since the evaluation of a  \SIF\ condition may trigger a 
  collection, after evaluating the condition   the available memory
  is  checked once  again against  a  revised estimate  of the  memory
  (based on  value of the condition)  required to execute the  rest of
  the program.  A GC is triggered if enough memory is unavailable.
\end{enumerate}

>>>>

\begin{figure}[t]\footnotesize
\renewcommand{\arraystretch}{0.9}
\begin{eqnarray*}
   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\,\,\, e_\mainpgm
   \hspace{5em} \,\,\,\,\,\,\,\,\,\; \mbox{\em --- program}\\
    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
\,\,x_n)\,\,
    e)
    \hspace{0.2em} \ \ \ \ \ \ \ \ \  \mbox{\em --- function def} \\
e \in \mathit{Expr} & \!\!\!::=\!\!\! &
\left\{\begin{array}{@{}ll@{\hspace{1em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
binding} \\
       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{App} & \!\!\!::=\!\!\!  &
\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
       (\CAR\,\, x) &  \mbox{\em --- selects $1^{st}$ part of \CONS} \\
       (\CDR\,\, x) & \mbox{\em --- selects $2^{nd}$ part of \CONS} \\
       (\NULLQ\,\, x) & \mbox{\em --- returns 0 if x is not $\NIL$}\\
       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  generic arithmetic} \\
%%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
    \end{array}\right.
\end{eqnarray*}
\vspace*{-6pt}
  \caption{The syntax of our language}\label{fig:lang-syntax}
%\figrule
\normalsize
\vspace*{-6pt}
\end{figure}

@@

Closure  variables  get  their  liveness values  through  a  chain  of
dependences beginning  at a  variable at an  evaluation point.   As an
example,  in  Figure~\ref{fig:lv-closure},   a  dependence  chain  for
$\px_{\mathit{\pi}_1}$  begins   with  the   variable  $\pz$   at  the
evaluation  point $\psi_3$.   The variable  \pz\ returned  at $\psi_3$
depends on  \py\ through  the expression  $(+~\py~1)$.  $\py$  in turn
depends  on   the  closure  variable   $\px_{\mathit{\pi}_1}$  through
$(\length~\px_{\mathit{\pi}_1})$.  We
denote this  chain of  dependences as  $\lbrack \psi_3\!\!\!:\!\!\!\pz
\leftarrow            (+~\py~1),\,           \py            \leftarrow
(\length~\px_{\mathit{\pi}_1})\rbrack$.    Indeed,   the   chains   of
closures  in   the  heap  are    runtime  representations   of  these
dependences.   Since  $\pz$  is  evaluated  at  $\psi_3$  due  to  the
expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
$\px_{\pi_1}$.  Other dependence chains which result in a liveness for
$\px_{\pi_1}$    are    $\lbrack    \psi_1\!\!\!:\!\!\py    \leftarrow
(\length~\px_{\pi_1})\rbrack$   and  $\lbrack   \psi_2\!\!\!:\!\!\!\pw
\leftarrow          (/~\pu~\py),\,\,           \py          \leftarrow
(\length~\px_{\pi_1})\rbrack$.   The  liveness analysis  described  in
this section declares  the liveness of $\px_{\pi_1}$ to be  a union of
the liveness arising  from these dependence chains.  To be  safe, a GC
during evaluation  of $\py$ at  $\psi_1$ has  to use this  liveness to
copy the heap starting from  $\px_{\pi_1}$.  However, notice that if a
GC  takes place  while evaluating  $\pz$  at $\psi_3$,  it can  safely
consider only the liveness arising from the dependence chain\linebreak
$\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow (+~\py~1),\, \py \leftarrow
(\length~\px_{\pi_1})\rbrack$.    The    garbage   collection   scheme
described in Section~\ref{sec:GC-scheme} uses a generalization of this
observation  to  dynamically  select   an  evaluation  point  specific
liveness in order to collect more garbage.

@@

Based on  the above considerations,  we restrict the  possible garbage
collection points in a function body to the following:
\begin{enumerate}
\item We  statically over-approximate the memory required  to create
  the closures for each function body.  On entering a function, if the
  available memory is  less than this requirement for  the function, a
  GC is triggered.
\item Since the evaluation of a  \SIF\ condition may trigger a 
  collection, after evaluating the condition   the available memory
  is  checked once  again against  a  revised estimate  of the  memory
  (based on  value of the condition)  required to execute the  rest of
  the program.  A GC is triggered if enough memory is unavailable.
\end{enumerate}
